\bf Calendar-based graphics for visualizing people's daily schedules

### 2019-05-08 main.Rmd
done with revision

We develop a method for organizing and visualizing temporal data, collected at sub-daily intervals, into a calendar layout. The calendar format is created using linear algebra, giving a restructuring of the data, that can then be integrated into a data pipeline. The core component of the pipeline is to visualise the resulting data using the grammar of graphics [@wilkinson2006grammar; @wickham2009ggplot2], as used in **ggplot2** [@R-ggplot2], which defines plots as a functional mapping from data points to graphical primitives. The data restructuring approach is consistent with the tidy data principles available in the **tidyverse** [@R-tidyverse] suite. The methods are implemented in a new R package called **sugrrants** [@R-sugrrants].

@@

In contrast, calendar-based graphics unpack the temporal variable, at different resolutions, to digest multiple seasonalities and special events. There is some existing work in this area. For example, @VanWijkCluster1999 developed a calendar view of the heatmap to represent the number of employees in the work place over a year, where colors indicate different clusters derived from the days. It contrasts week days and weekends, highlights public holidays, and presents other known seasonal variation such as school vacations, all of which have influence over the turn-outs in the office. Alongside @Jones-heatmap, @R-TimeProjection, @R-ggTimeSeries, and @R-ggcal implemented some variants of calendar-based heatmaps as in R packages: **TimeProjection**, **ggTimeSeries**, and **ggcal** respectively. However, these techniques are limited to color-encoding graphics and are unable to use time scales smaller than a day. Time of day, which serves as one of the most important aspects in explaining substantial variations arising from the pedestrian sensor data, will be neglected through daily aggregation. Additionally, if simply using colored blocks rather than curves, it may become perceptually difficult to estimate the shape positions and changes, although using curves comes with the cost of more display capacity [@cleveland1984graphical; @lam2007overview].

@@

The remainder of the paper is organized as follows. Section \@ref(sec:algorithm) demonstrates the construction of the calendar layout in depth. Section \@ref(sec:transformation) describes the algorithms of data transformation. Section \@ref(sec:opt) lists and describes the options that come with the `frame_calendar()` function. Section \@ref(sec:variations) presents some variations of its usage. The calendar layout is a type of faceting, and Section \@ref(sec:facet-calendar) develops this into a fully-fledged faceting method (`facet_calendar()`) with formal labels and axes. Graphical analyses of sub-daily people's activities are illustrated with a case study in Section \@ref(sec:case). Section \@ref(sec:discussion) discusses the limitations of calendar displays and possible new directions.

@@

The use of the calendar display is illustrated on smart meter energy usage from four households in Melbourne, Australia. Individuals can download their own data from the energy supplier, and the data contained in the paper is made available from four colleagues of the authors. The calendar display is useful to help people understand their energy use. The data contains half-hourly electricity consumption in the first half of 2018. The analysis begins by looking at the distribution over days of week, then time of day split by work days and non-work days, followed by the calendar display to inspect the daily schedules.

@@

>>>>

We develop a method for organizing and visualizing temporal data, collected at sub-daily intervals, into a calendar layout. The calendar format is created using linear algebra, giving a restructuring of the data, that can then be integrated into a data pipeline. The core component of the pipeline is to visualise the resulting data using the grammar of graphics [@wilkinson2006grammar; @wickham2009ggplot2], as used in **ggplot2** [@R-ggplot2], where plots are defined as a functional mapping from data points to graphical primitives. The data restructuring approach is consistent with the tidy data principles available in the **tidyverse** [@R-tidyverse] suite. The methods are implemented in a new R package called **sugrrants** [@R-sugrrants].

@@

In contrast, calendar-based graphics unpack the temporal variable, at different resolutions, to digest multiple seasonalities and special events. There are some existing works in this area. For example, @VanWijkCluster1999 developed a calendar view of the heatmap to represent the number of employees in the work place over a year, where colors indicate different clusters derived from the days. It contrasts week days and weekends, highlights public holidays, and presents other known seasonal variation such as school vacations, all of which have influence over the turn-outs in the office. Alongside @Jones-heatmap, @R-TimeProjection, @R-ggTimeSeries, and @R-ggcal implemented some variants of calendar-based heatmaps as in R packages: **TimeProjection**, **ggTimeSeries**, and **ggcal** respectively. However, these techniques are limited to color-encoding graphics and are unable to use time scales smaller than a day. Time of day, which serves as one of the most important aspects in explaining substantial variations arising from the pedestrian sensor data, will be neglected through daily aggregation. Additionally, if simply using colored blocks rather than curves, it may become perceptually difficult to estimate the shape positions and changes, although using curves comes with the cost of more display capacity [@cleveland1984graphical; @lam2007overview].

@@

The remainder of the paper is organized as follows. Section \@ref(sec:algorithm) demonstrates the construction of the calendar layout in depth. Section \@ref(sec:transformation) describes the algorithms of data transformation. Section \@ref(sec:opt) lists and describes the options that come with the `frame_calendar()` function. Section \@ref(sec:variations) presents some variations of its usage. The calendar layout is a type of faceting, and Section \@ref(sec:facet-calendar) extends this into a fully-fledged faceting method as in `facet_calendar()` with formal labels and axes. Graphical analyses of sub-daily people's activities are illustrated with a case study in Section \@ref(sec:case). Section \@ref(sec:discussion) discusses the limitations of calendar displays and possible new directions.

@@

The use of the calendar display is illustrated on smart meter energy usage from four households in Melbourne, Australia. Individuals can download their own data from the energy supplier, and the data used in this section is sourced from four colleagues of the authors. The calendar display is useful to help people understand their energy use. The data contains half-hourly electricity consumption in the first half of 2018. The analysis begins by looking at the distribution over days of week, then time of day split by work days and non-work days, followed by the calendar display to inspect the daily schedules.

@@

```{r write-bib}
write_bib(c("ggplot2", "tidyverse", "sugrrants", "geofacet", "shiny", "wanderer4melb"), 'packages.bib')
```

### 2019-05-08 main.tex
done with revision

We develop a method for organizing and visualizing temporal data, collected at sub-daily intervals, into a calendar layout. The calendar format is created using linear algebra, giving a restructuring of the data, that can then be integrated into a data pipeline. The core component of the pipeline is to visualise the resulting data using the grammar of graphics \citep{wilkinson2006grammar, wickham2009ggplot2}, as used in \textbf{ggplot2} \citep{R-ggplot2}, which defines plots as a functional mapping from data points to graphical primitives. The data restructuring approach is consistent with the tidy data principles available in the \textbf{tidyverse} \citep{R-tidyverse} suite. The methods are implemented in a new R package called \textbf{sugrrants} \citep{R-sugrrants}.

@@

In contrast, calendar-based graphics unpack the temporal variable, at different resolutions, to digest multiple seasonalities and special events. There is some existing work in this area. For example, \citet{VanWijkCluster1999} developed a calendar view of the heatmap to represent the number of employees in the work place over a year, where colors indicate different clusters derived from the days. It contrasts week days and weekends, highlights public holidays, and presents other known seasonal variation such as school vacations, all of which have influence over the turn-outs in the office. Alongside \citet{Jones-heatmap}, \citet{R-TimeProjection}, \citet{R-ggTimeSeries}, and \citet{R-ggcal} implemented some variants of calendar-based heatmaps as in R packages: \textbf{TimeProjection}, \textbf{ggTimeSeries}, and \textbf{ggcal} respectively. However, these techniques are limited to color-encoding graphics and are unable to use time scales smaller than a day. Time of day, which serves as one of the most important aspects in explaining substantial variations arising from the pedestrian sensor data, will be neglected through daily aggregation. Additionally, if simply using colored blocks rather than curves, it may become perceptually difficult to estimate the shape positions and changes, although using curves comes with the cost of more display capacity \citep{cleveland1984graphical, lam2007overview}.

@@

The remainder of the paper is organized as follows. Section \ref{sec:algorithm} demonstrates the construction of the calendar layout in depth. Section \ref{sec:transformation} describes the algorithms of data transformation. Section \ref{sec:opt} lists and describes the options that come with the \texttt{frame\_calendar()} function. Section \ref{sec:variations} presents some variations of its usage. The calendar layout is a type of faceting, and Section \ref{sec:facet-calendar} develops this into a fully-fledged faceting method (\texttt{facet\_calendar()}) with formal labels and axes. Graphical analyses of sub-daily people's activities are illustrated with a case study in Section \ref{sec:case}. Section \ref{sec:discussion} discusses the limitations of calendar displays and possible new directions.

>>>>

We develop a method for organizing and visualizing temporal data, collected at sub-daily intervals, into a calendar layout. The calendar format is created using linear algebra, giving a restructuring of the data, that can then be integrated into a data pipeline. The core component of the pipeline is to visualise the resulting data using the grammar of graphics \citep{wilkinson2006grammar, wickham2009ggplot2}, as used in \textbf{ggplot2} \citep{R-ggplot2}, where plots are defined as a functional mapping from data points to graphical primitives. The data restructuring approach is consistent with the tidy data principles available in the \textbf{tidyverse} \citep{R-tidyverse} suite. The methods are implemented in a new R package called \textbf{sugrrants} \citep{R-sugrrants}.

@@

In contrast, calendar-based graphics unpack the temporal variable, at different resolutions, to digest multiple seasonalities and special events. There are some existing works in this area. For example, \citet{VanWijkCluster1999} developed a calendar view of the heatmap to represent the number of employees in the work place over a year, where colors indicate different clusters derived from the days. It contrasts week days and weekends, highlights public holidays, and presents other known seasonal variation such as school vacations, all of which have influence over the turn-outs in the office. Alongside \citet{Jones-heatmap}, \citet{R-TimeProjection}, \citet{R-ggTimeSeries}, and \citet{R-ggcal} implemented some variants of calendar-based heatmaps as in R packages: \textbf{TimeProjection}, \textbf{ggTimeSeries}, and \textbf{ggcal} respectively. However, these techniques are limited to color-encoding graphics and are unable to use time scales smaller than a day. Time of day, which serves as one of the most important aspects in explaining substantial variations arising from the pedestrian sensor data, will be neglected through daily aggregation. Additionally, if simply using colored blocks rather than curves, it may become perceptually difficult to estimate the shape positions and changes, although using curves comes with the cost of more display capacity \citep{cleveland1984graphical, lam2007overview}.

@@

The remainder of the paper is organized as follows. Section \ref{sec:algorithm} demonstrates the construction of the calendar layout in depth. Section \ref{sec:transformation} describes the algorithms of data transformation. Section \ref{sec:opt} lists and describes the options that come with the \texttt{frame\_calendar()} function. Section \ref{sec:variations} presents some variations of its usage. The calendar layout is a type of faceting, and Section \ref{sec:facet-calendar} extends this into a fully-fledged faceting method as in \texttt{facet\_calendar()} with formal labels and axes. Graphical analyses of sub-daily people's activities are illustrated with a case study in Section \ref{sec:case}. Section \ref{sec:discussion} discusses the limitations of calendar displays and possible new directions.
